Deep Neural Architectures for Discourse Segmentation in E-Mail Based Behavioral Interventions

### 2018-11-26 paper-amia-summit-2019/manuscript.tex
final version by Alex

\author{Mehedi Hasan, BS$^{1}$\footnote{Authors provided an equal contribution. \label{footnote1}}, Alexander Kotov, PhD$^{1}$\textsuperscript{\ref{footnote1}}, Sylvie Naar, PhD$^{2}$, Gwen L. Alexander, PhD$^{3}$, April Idalski Carcone, PhD$^{4}$}

@@

\noindent{\bf Abstract}
\textit{Communication science approaches to develop effective behavior interventions, such as motivational interviewing (MI), are limited by traditional qualitative coding of communication exchanges, a very resource-intensive and time-consuming process. This study focuses on the analysis of e-Coaching sessions, behavior interventions delivered via email and grounded in the principles of MI. A critical step towards automated qualitative coding of e-Coaching sessions is segmentation of emails into fragments that correspond to MI behaviors. This study frames email segmentation task as a classification problem and utilizes word and punctuation mark embeddings in conjunction with part-of-speech features to address it. We evaluated the performance of conditional random fields (CRF) as well as multi-layer perceptron (MLP), bi-directional recurrent neural network (BRNN) and convolutional recurrent neural network (CRNN) for the task of email segmentation. Our results indicate that CRNN outperforms CRF, MLP and BRNN achieving 0.989 macro F1-score overall and 0.825 macro F1-score for new segment detection.}

\section*{Introduction}
The emergence of e-Health technologies has greatly expanded the reach of behavioral interventions. One such intervention is email-delivered Motivational Interviewing (MI). MI is an evidence-based communication technique to increase intrinsic motivation and self-efficacy for behavior change\cite{miller2012motivational,miller2009toward}. MI is linked to behavior change through the elicitation of patient ``change talk'', or statements of intrinsic motivation expressing patients’ own desires, abilities, reasons, need for, and commitment to behavior change\cite{apodaca2009mechanisms}. However, communication science approaches to understanding the efficacy of MI are inherently limited by traditional qualitative coding methods. The specific provider behaviors responsible for the elicitation of change talk, however, are less clear and may vary by treatment context. Thus, a current focus of MI research is understanding which provider behaviors in which contexts lead to patient change talk. To date, no study has examined this process in email-delivered MI.

@@

In this study, we focus on the analysis of email-delivered MI, or e-Coaching, to promote healthy eating among young adults. The e-Coaching dataset is composed of email correspondence between a MI counselor and the young adult patient. Unlike transcribed in-person discourse, email correspondence is not clearly segmented into codable speech acts (i.e., utterances). Thus, the unstructured nature of e-Coaching exchanges poses a unique set of analytic challenges. Segmentation of e-Coaching exchanges into textual fragments that correspond to distinct e-Coach and patient communication behaviors is a significant barrier to qualitative analysis of this type of clinical conversation. Automating this task is a unique and challenging problem due to the following reasons:

\begin{enumerate}
\item Emails are unstructured text containing informal information exchange in a non-traditional format. For example, an e-Coach usually responds to several previous patient statements in one email. In contrast, in a traditional, in-person MI session, an utterance is assumed to be a response to the immediately preceding utterance.
\item Discourse segments in e-Coaching do not have a clear breakpoint, such as the end of a sentence or paragraph. One sentence might be divided into fragments corresponding to multiple MI behaviors. On the other hand, an MI behavior may comprise several sentences.
\end{enumerate}

@@

Figure~\ref{fig:text-segment} illustrates a segmentation of an e-Coaching exchange, in which the first sentence is segmented into 2 MI behavior fragments, while the fourth and fifth segments correspond to one and three sentences, respectively. Segmentation of e-Coaching exchanges constitutes a special case of clinical discourse analysis \cite{webber2012discourse} aimed at better understanding the effective communication strategies specific to this type of behavioral interventions.

The goal of this study is to assess the effectiveness of deep learning methods for the task of automated segmentation of e-Coaching emails into textual fragments corresponding to individual patient and provider behaviors.  For this study, we utilized the data from MENU GenY (Making Effective Nutrition Choices for Generation Y) \cite{alexander2017motivations}, a web-delivered public health intervention with email-based coaching to encourage increased fruit and vegetable intake among young adults, aged 21–-30. A secondary goal of the MENU GenY project was to identify the specific communication strategies used by e-Coaches to elicit change talk for healthier eating among young adult patients. Segmentation of clinical conversation in the context of electronically delivered interventions into groups of MI behaviors, is traditionally performed manually by MI researchers, which significantly slows down their qualitative analysis. This paper is the first work to evaluate the empirical effectiveness of deep learning architectures in addressing \textit{the problem of discourse segmentation in the context of e-mail-based behavioral interventions}.

Specifically, we introduce and evaluate the effectiveness of distributed word representations (i.e. word embeddings) as well as lexical, punctuation and part-of-speech (POS) features in conjunction with both traditional supervised machine learning methods, such as linear-chain Conditional Random Fields (CRF)\cite{lafferty2001conditional} and deep learning methods, such as Multi-Layer Perceptron (MLP),\cite{rumelhart1986learning} Bidirectional Recurrent Neural Network (BRNN)\cite{schuster1997bidirectional} and Convolutional Recurrent Neural Network (CRNN),\cite{treviso2017sentence} to find the best performing method and feature combination.

@@

Prior work on textual segmentation in the biomedical domain primarily focused on sentence boundary detection \cite{griffis2016quantitative,kreuzthaler2015detection,treviso2017sentence} and segmentation of clinical documents in patients' electronic health records (EHR) into sections and headers. \cite{apostolova2009automatic,denny2009evaluation,tepper2012statistical,cho2002text} In particular, Maximum Entropy models \cite{tepper2012statistical} and Support Vector Machine (SVM) along with word-vector cosine similarity metric and several heuristics \cite{apostolova2009automatic} have been applied to identify specific sections in EHR, such as general patient information, medical history, procedures, findings, etc. Denny et al. \cite{denny2009evaluation} proposed SecTag algorithm, which combined natural language processing techniques, terminology-based rules and a Na\"{i}ve Bayes classifier to identify sections and headers in EHR. Segmentation of e-Coaching emails, however, is different from segmentation of other clinical documents, since the focus is on dialog acts in clinical conversation.

SVM in conjunction with prosodic and part-of-speech features \cite{kreuzthaler2015detection} and Recurrent Convolutional Neural Networks \cite{griffis2016quantitative} have also been utilized for \textit{sentence boundary detection} in general text. Liu et al. \cite{liu2005using} demonstrated that a linear-chain CRF outperforms Hidden Markov and Maximum Entropy models for this task.

Segmentation of e-Coaching emails is also different from traditional shallow discourse analysis \cite{galley2003discourse}, which besides identification of speech acts, also aims to determine the types of transitions between speech acts and labels speech acts with speakers who performed them in a multi-speaker conversation. The proposed methods will automate the process of segmenting clinical exchanges into MI behaviors, which will significantly reduce the time and resources required to perform such segmentation manually. Furthermore, these methods can be integrated with auto coding methods \cite{hasan2016study,kotov2015interpretable} to create a software pipeline for fully automated analysis of email-delivered behavioral interventions.

\section*{Materials and methods}
\subsection*{\textit{Data collection}}
The experimental dataset for this work was constructed from 49 e-Coaching sessions, which include 330 and 281 emails by e-Coaches and patients, respectively. Various statistics of experimental dataset are provided in Table~\ref{tab:datastat}. Each session represents an MI intervention delivered via email. Emails were segmented into 3,138 text fragments and and annotated using the Minority Youth-Sequential Coding of Process Exchanges (MYSCOPE),~\cite{carcone2013provider} a qualitative coding scheme to characterize patient-provider communication during MI sessions with 115 distinct behavior codes. We consider email segmentation as a binary classification problem, in which each word or punctuation mark is annotated with a class label ``new segment'' or ``same segment'' to indicate whether it precedes a new segment or not. In total, the dataset consists of 95,777 words and 7,140 punctuation marks and includes 3,138 ``new segment'' and 99,779 ``same segment'' instances. In this study, we experimented with traditional machine learning methods, such as Conditional Random Fields (CRF)\cite{lafferty2001conditional} and deep learning methods, such as Multi-Layer Perceptron (MLP),\cite{rumelhart1986learning} Bidirectional Recurrent Neural Network (bidirectional RNN or BRNN)\cite{schuster1997bidirectional} and Convolutional Recurrent Neural Network (CRNN).\cite{treviso2017sentence} For an MLP model, training and testing samples were created based on a sliding window of $2n$ words or punctuation marks over each position (which could be a word or a punctuation mark) in a given input sequence, such that each instance consists of the $n$ words or punctuation marks after this position and $n$ words or punctuation marks prior to this position, including the position itself. Each sample is assigned a ``new segment'' or ``same segment'' label based on whether there should be a segment break after the current position or not. In the case of CRF, BRNN and CRNN models, an e-Coaching email was taken as an input sequence, POS tags and word embeddings of each word or punctuation mark were used as input and binary labels corresponding to ``new segment'' and ``same segment'' classification decisions were considered as the model output. In the gold standard, words or punctuations within the same segment were assigned the label of 0 and the last word or punctuation mark of a segment were assigned the label of 1 (an example of a sequence with assigned labels is provided in the $2^{nd}$ row of Table~\ref{tab:datastat}). \\

\begin{table}[ht]
\centering
\caption{\textbf{Summary of statistics of experimental dataset.}}
\label{tab:datastat}
 \begin{tabular}{|l|l|l|l|l|l|l|l|l|l|}
  \hline
   \multirow{2}{*}{\textbf{Sessions}} & \multirow{2}{*}{\textbf{Instances}} & \multicolumn{2}{|c|}{\textbf{Class Segments}} & \multicolumn{2}{|c|}{\textbf{Tokens}} & \multicolumn{2}{|c|}{\textbf{Emails}} & \multicolumn{2}{|c|}{\textbf{Annotation}} \\\cline{3-10}
   &  & \textbf{new}  & \textbf{same} & \textbf{words} & \textbf{punctuations}  & \textbf{patient} & \textbf{provider} & \textbf{method}  & \textbf{codes} \\ \hline    
 49 & 102,917 & 3,138 & 99,779 & 95,777 & 7,140 & 281 & 330 & MYSCOPE & 115 \\ \hline
 \multicolumn{10}{|c|}{segmentation example: Hi[0] XXX[0] ,[0] its[0] good[0] to[0] hear[0] from[0] you[0] .[1] it[0] sounds[0] like[0]...} \\ \hline
  \end{tabular}
\end{table}

\subsection*{\textit{Features}}
We utilized three types of features in conjunction with CRF, MLP, BRNN and CRNN methods: word embeddings as lexical features, punctuation and POS features. Syntactic abstractions of individual words, such as POS tags, have been previously shown to be effective features for similar natural language processing tasks.\cite{liu2005using,treviso2017sentence} To extract POS features, we pre-processed e-Coaching emails using the NLTK POS tagger\footnote{\url{https://www.nltk.org/}}. Punctuation marks, which correspond to one of the symbols \{`.', `,', `!', `?', `:', `;'\} between a pair of words, were also used as a feature, since punctuation marks designate the boundary of a sentence, clause or phrase and often also correspond to a segment boundary.\cite{cho2002text} For natural language processing (NLP) tasks, inputs are received as a text, in which individual words are as the basic lexico-semantic units. Therefore, it is important to represent a word in such a way that preserves all relevant lexical and semantic information. Dense real-valued vector (i.e. embedding) is a form of distributed word representation, in which each word is associated with a vector in a low-dimensional vector space. Embeddings have been previously shown to effectively capture semantic, syntactic and morphological properties of words.\cite{pennington2014glove, mikolov2013distributed} For experiments reported in this paper, we utilized word embeddings pre-trained on Google News corpus consisting of 1.6 billion words using word2vec software package.\footnote{\label{fn:word2vec}https://code.google.com/p/word2vec/} For words or punctuation marks, which do not have pre-trained embeddings, we utilized the embeddings of the same dimensionality trained on the experimental dataset. CRF utilized lexical features, POS tags and the preceding label.

@@

\textbf{Conditional Random Field}: CRF has been widely used in various sequence annotation NLP tasks, such as part-of-speech tagging.\cite{lafferty2001conditional, hirohata2008identifying} Unlike a maximum entropy Markov model, which uses per-state exponential models for conditional probability of the next state given a current state, CRF model directly estimates a distribution of the entire output sequence conditioned on the observation sequence. A traditional linear-chain CRF model is defined as a conditional probability distribution $p(y|x)$ of output sequence $y$, given input sequence $x$:

@@

where $Z_x$ is a normalization factor, $f_k(y_{t-1}, y_t, x, t)$ is a feature function, and $\lambda_k$ is a learned weight associated with feature $f_k$. The optimal output sequence $y^*$ for input sequence $x$, $y^* = {arg\,max}_y p(y|x)$, is obtained efficiently using the Viterbi algorithm. In our experiments, the following features were utilized in conjunction with CRF models: i) current word or punctuation ii) next and previous 3 words or punctuations iii) binary feature indicating whether a word or punctuation is a special character (';', '?', '.', ',', '!', ':', etc.) or not iv) binary feature indicating whether a word is a title word or not (e.g. ``The'' is a title word but ``the'' is not) v) POS tags.

\textbf{Multi-Layer Perceptron}: MLP is a neural network, which consists of multiple fully connected layers that map an input onto one or several outputs.\cite{rumelhart1986learning} Figure~\ref{fig:mlp} illustrates a multi-layer perceptron with a single hidden layer. MLPs have no cycles or loops. Information in them flows only forward, from the input layer through the hidden layer(s) to the output layer. The MLP in this study utilizes one hidden layer consisting of 128 neurons and rectified linear unit (ReLU) as a nonlinear activation function. In order to prevent over-fitting, we applied dropout (random masking of neurons) \cite{srivastava2014dropout} to fully connected layers during training. Dropout was also applied to fully connected layers in BRNN and CRNN.

@@

\textbf{Bidirectional Recurrent Neural Network}: BRNN is a neural network designed to capture sequential patterns by considering both past  and future inputs as well as complex relationships between input features and output labels.\cite{schuster1997bidirectional} The hidden state of BRNN is an aggregation of the hidden states of a forward and backward recurrent neural networks (RNNs). Gated Recurrent Units (GRU)\cite{chung2014empirical} capable of handling variable size input sequence and having internal memory, which can be reset, were utilized as an RNN. %Figure~\ref{fig:crnn} represents the architecture of BRNN, in the case when a convolution layer is removed.

\textbf{Convolutional Recurrent Neural Network}: CRNN \cite{treviso2017sentence} shown in Figure~\ref{fig:crnn} is a deep neural network architecture, which combines convolutional and recurrent layers. Our implementation of CRNN consists of 5 layers: 1) input layer 2) embedding layer 3) convolution layer with max pooling 4) BRNN layer 5) fully connected layer with dropout and sigmoid output. E-coaching email exchanges are represented as a sequence of $m$ words and punctuations, which are fed into the input and embedding layers to produce a $m \times n_e$ matrix after fetching the embeddings for words and punctuations in the input sequence. This matrix corresponds to the distributed representation of an input email exchange, which contains rich morpho-syntactic information that can be utilized for its segmentation. When POS tags are utilized along with word embeddings, they are represented with a 10-dimensional vector, which is concatenated with 300-dimensional word embeddings to obtain new embedding vectors $n_e = [n_w;n_p]$ of size 310. The primary purpose of a convolution layer is to extract new features for each word or punctuation mark based on the neighboring words or punctuation marks. A one-dimensional (1D) convolution operation is utilized in this layer in our implementation of BRNN. In 1D convolution, one filter is responsible for extraction of one feature. After applying $n_f$ different filters with zero-padding on both sides of the input text, $n_f$ features are produced by the convolution layer for each word. A max pooling over time operation was then performed to find the most significant features in a textual fragment. The bidirectional recurrent layer receives new features extracted by the convolution layer. Unidirectional RNNs are usually used to capture long-range dependencies in a sequence of observations. Bidirectional RNNs, on the other hand, are capable of capturing both past and future contexts through forward and backward traversals of a sequence. The purpose of the fully connected layer is to use the output of the bidirectional RNN layer for classifying each word or punctuation into ``new segment'' or ``same segment'' classes. Since a fully connected layer has a larger number of parameters, they are more likely to excessively co-adapt to  other parameters, which may result in over-fitting. To prevent this, we utilized dropout by randomly ignoring 50\% of the connections in the fully connected layer. Finally, logistic sigmoid outputs the probability of classifying or labeling each word or punctuation mark with ``same segment'' class. We experimentally determined the optimal parameters using 5-fold cross-validation and found out that the best performance is achieved when filter length in the convolution layer is 7, number of filters is 100, max-pool size is 3, ReLU is used as an activation function in the convolution layer, hyperbolic tangent is used as an activation function in the bi-directional RNN layer and the number of dimensions in the hidden state of RNNs is 200. Adam\cite{kingma2014adam} with 50 epochs, the batch size of 32 and learning rate of 0.001 was used for optimization and the early stopping strategy was applied.\footnote{source code of all methods is available at \url{https://github.com/teanalab/eCoaching-Text-Segmentation}}

\begin{figure}[!htb]
    \centering
    \includegraphics[width=0.95\textwidth]{figures/CRNN.eps}
    \caption{\textbf{Architecture of a convolutional recurrent neural network for automated segmentation of e-Coaching emails}}
    \label{fig:crnn}
\end{figure}

\subsection*{\textit{Evaluation metrics}}
We report standard metrics of precision, recall and F1-measure to evaluate the performance of the classifiers.\cite{aas1999text} Accuracy is not reported as a performance metric, since it is highly sensitive to the prior class probabilities and does not fully describe the actual difficulty of the decision problem, when highly unbalanced datasets are involved. The results are reported based on 5-fold cross-validation (one fold was used as a test set and the remaining 4 folds were utilized as a training set) and weighted macro-averaging over the folds. We also report the area under the precision-recall curve (AUPR), due to its effectiveness in measuring the performance of binary classifiers in the case of the datasets with imbalanced class distribution.\cite{davis2006relationship}

\section*{Results}
Our experiments spanned three dimensions. First, we determined the optimal size of word embedding vectors and the sliding window of MLP. Second, we evaluated the performance of different methods with respect to detecting ``new segment'' as well as the weighted average over ``new segment'' and ``same segment'' classes. Third, we assessed the impact of different types of features as well as their combination on the performance of different machine learning methods on the e-Coaching email segmentation task.

>>>>

\author{Mehedi Hasan, BS$^{1}$\footnote{These authors provided an equal contribution. \label{footnote1}}, Alexander Kotov, PhD$^{1}$\textsuperscript{\ref{footnote1}}, Sylvie Naar, PhD$^{2}$, Gwen L. Alexander, PhD$^{3}$, April Idalski Carcone, PhD$^{4}$}

@@

\noindent{\bf Abstract}
\textit{Communication science approaches to develop effective behavior interventions, such as motivational interviewing (MI), are limited by traditional qualitative coding of communication exchanges, a very resource-intensive and time-consuming process. This study focuses on the analysis of e-Coaching sessions, behavior interventions delivered via email and grounded in the principles of MI. A critical step towards automated qualitative coding of e-Coaching sessions is segmentation of emails into fragments that correspond to MI behaviors. This study frames email segmentation task as a classification problem and utilizes word and punctuation mark embeddings in conjunction with part-of-speech features to address it. We evaluated the performance of conditional random fields (CRF) as well as multi-layer perceptron (MLP), bi-directional recurrent neural network (BRNN) and convolutional recurrent neural network (CRNN) for the task of email segmentation. Our results indicate that CRNN outperforms CRF, MLP and BRNN achieving 0.989 weighted macro-averaged F1-measure and 0.825 F1-measure for new segment detection.}

\section*{Introduction}
The emergence of e-Health technologies has greatly expanded the reach of behavioral interventions. One such intervention is email-delivered Motivational Interviewing (MI). MI is an evidence-based communication technique to increase intrinsic motivation and self-efficacy for behavior change\cite{miller2012motivational,miller2009toward}. MI is linked to behavior change through the elicitation of patient ``change talk'', or statements of intrinsic motivation expressing patients’ own desires, abilities, reasons, need for, and commitment to behavior change\cite{apodaca2009mechanisms}. However, communication science approaches to understanding the efficacy of MI are inherently limited by traditional qualitative coding methods. The specific provider behaviors responsible for the elicitation of change talk, however, are less clear and may vary by treatment context. Thus, current MI research is focused on understanding which provider behaviors and in which contexts lead to patient change talk. To date, no study has examined this process in email-delivered MI.

@@

In this study, we focus on the analysis of email-delivered MI, or e-Coaching, to promote healthy eating among young adults. The e-Coaching dataset is composed of email correspondence between a MI counselor and the young adult patient. Unlike transcribed in-person exchanges, email correspondence is not clearly segmented into codable speech acts (i.e., utterances). Thus, the unstructured nature of e-Coaching exchanges poses a unique set of analytic challenges. Segmentation of e-Coaching exchanges into textual fragments that correspond to distinct e-Coach and patient communication behaviors is a significant barrier to qualitative analysis of this type of clinical conversation. Automating this task is a unique and challenging problem due to the following reasons:

\begin{enumerate}
\item Emails are unstructured text containing informal information exchange in a non-traditional format. For example, an e-Coach usually responds to several previous patient statements in one email. In contrast, in a traditional, in-person MI session, each utterance is assumed to be a response to an immediately preceding utterance.
\item Discourse segments in e-Coaching do not have a clear breakpoint, such as the end of a sentence or a paragraph. One sentence may be divided into fragments corresponding to multiple MI behaviors. On the other hand, an MI behavior may comprise several sentences.
\end{enumerate}

@@

Figure~\ref{fig:text-segment} illustrates a segmentation of an e-Coaching exchange, in which the first sentence is segmented into 2 MI behavior fragments, while the fourth and fifth MI behavior fragments comprise one and three sentences, respectively. Segmentation of e-Coaching exchanges constitutes a special case of clinical discourse analysis \cite{webber2012discourse} aimed at better understanding the effective communication strategies specific to this type of behavioral interventions.

The goal of this study is to assess the effectiveness of deep learning methods for the task of automated segmentation of e-Coaching emails into textual fragments corresponding to individual patient and provider behaviors.  For this study, we utilized the data from MENU GenY (Making Effective Nutrition Choices for Generation Y) \cite{alexander2017motivations}, a web-delivered public health intervention with email-based coaching to encourage increased fruit and vegetable intake among young adults, aged 21–30. A secondary goal of the MENU GenY project was to identify the specific communication strategies used by e-Coaches to elicit change talk for healthier eating among young adult patients. Segmentation of clinical conversation in the context of electronically delivered interventions into groups of MI behaviors is traditionally performed manually by MI researchers, which significantly slows down its qualitative analysis. \textit{This paper is the first work to evaluate the empirical effectiveness of deep learning architectures in addressing the problem of discourse segmentation in the context of email-based behavioral interventions}.

Specifically, we evaluate the effectiveness of distributed representations (i.e. embeddings) of words and punctuation marks as well as part-of-speech (POS) features in conjunction with both traditional supervised machine learning methods, such as linear-chain Conditional Random Fields (CRF)\cite{lafferty2001conditional} and deep learning methods, such as Multi-Layer Perceptron (MLP),\cite{rumelhart1986learning} Bidirectional Recurrent Neural Network (BRNN)\cite{schuster1997bidirectional} and Convolutional Recurrent Neural Network (CRNN),\cite{treviso2017sentence} to determine the best performing method and feature combination for the task of segmentation of e-Coaching emails into MI behaviors.

@@

Prior work on textual segmentation in the biomedical domain primarily focused on sentence boundary detection \cite{griffis2016quantitative,kreuzthaler2015detection,treviso2017sentence} and segmentation of clinical documents in patients' electronic health records (EHR) into sections and headers. \cite{apostolova2009automatic,denny2009evaluation,tepper2012statistical,cho2002text} In particular, maximum entropy models \cite{tepper2012statistical} and Support Vector Machine (SVM) along with word vector similarity metrics and several heuristics \cite{apostolova2009automatic} have been applied to identify specific sections in EHR, such as general patient information, medical history, procedures, findings, etc. Denny et al. \cite{denny2009evaluation} proposed SecTag algorithm, which combined natural language processing techniques, terminology-based rules and a Na\"{i}ve Bayes classifier to identify sections and headers in EHR. Segmentation of e-Coaching emails, however, is different from segmentation of other clinical documents, since the focus is on dialog acts in clinical conversation.

SVM in conjunction with prosodic and part-of-speech features \cite{kreuzthaler2015detection} and recurrent convolutional neural networks \cite{griffis2016quantitative} have also been utilized for \textit{sentence boundary detection} in general text. Liu et al. \cite{liu2005using} demonstrated that a linear-chain CRF outperforms Hidden Markov and maximum entropy models for this task.

Segmentation of e-Coaching emails is also different from traditional shallow discourse analysis \cite{galley2003discourse}, which besides identification of speech acts, also aims to determine the types of transitions between speech acts and label speech acts with the speakers who performed them in a multi-speaker conversation. The proposed methods will automate the process of segmenting clinical exchanges into MI behaviors, which will significantly reduce the time and resources required to perform such segmentation manually. Furthermore, these methods can be integrated with the automated MI behavior coding methods \cite{hasan2016study,kotov2015interpretable} to create a software pipeline for fully automated analysis of email-delivered behavioral interventions.

\section*{Materials and methods}
\subsection*{\textit{Dataset}}
The experimental dataset for this work was constructed from 49 e-Coaching sessions, which include 330 and 281 emails by e-Coaches and patients, respectively. Various statistics of the experimental dataset are provided in Table~\ref{tab:datastat}. Each e-Coaching session represents a MI intervention delivered via email. Emails were segmented into 3,138 text fragments and annotated with the Minority Youth-Sequential Coding of Process Exchanges (MY-SCOPE),~\cite{carcone2013provider} a qualitative coding scheme to characterize patient-provider communication during MI sessions with 115 distinct behavior codes. Email segmentation can be considered as sequence tagging, which can be framed as a binary classification problem, in which each word or punctuation mark is annotated with one of the two class labels (``new segment'' or ``same segment'') to indicate whether it is a beginning of a new MI behavior segment or not. In total, the dataset consists of 95,777 words and 7,140 punctuation marks and includes 3,138 ``new segment'' and 99,779 ``same segment'' instances. In this study, we experimented with traditional machine learning methods, such as Conditional Random Fields (CRF)\cite{lafferty2001conditional} and deep learning methods, such as Multi-Layer Perceptron (MLP),\cite{rumelhart1986learning} Bi-directional Recurrent Neural Network (BRNN)\cite{schuster1997bidirectional} and Convolutional Recurrent Neural Network (CRNN).\cite{treviso2017sentence} In the case of MLP, training and testing samples were created based on a sliding window of $2n$ words or punctuation marks over each position (which could be a word or a punctuation mark) in a given input sequence, such that each sample consists of the $n$ words or punctuation marks after the current position and $n$ words or punctuation marks prior to the current position, including the position itself. In the case of CRF, BRNN and CRNN models, an e-Coaching email was taken as an input sequence, POS tags and embeddings of each word or punctuation mark were used as input and binary labels corresponding to ``new segment'' and ``same segment'' classification decisions were considered as the model output. In the gold standard, words or punctuations within the same segment were assigned the label of 0 and the last word or punctuation mark of a segment were assigned the label of 1 (an example of an input with the assigned labels is provided in the $2^{nd}$ row of Table~\ref{tab:datastat}).

\begin{table}[ht]
\centering
\caption{\textbf{Summary of statistics of the experimental dataset and example of segmented sequence.}}
\label{tab:datastat}
 \begin{tabular}{|l|l|l|l|l|l|l|l|l|l|}
  \hline
   \multirow{2}{*}{\textbf{Sessions}} & \multirow{2}{*}{\textbf{Instances}} & \multicolumn{2}{|c|}{\textbf{Class labels}} & \multicolumn{2}{|c|}{\textbf{Tokens}} & \multicolumn{2}{|c|}{\textbf{Emails}} & \multicolumn{2}{|c|}{\textbf{Annotation}} \\\cline{3-10}
   &  & \textbf{new}  & \textbf{same} & \textbf{words} & \textbf{punctuations}  & \textbf{patient} & \textbf{provider} & \textbf{method}  & \textbf{codes} \\ \hline    
 49 & 102,917 & 3,138 & 99,779 & 95,777 & 7,140 & 281 & 330 & MYSCOPE & 115 \\ \hline
 \multicolumn{10}{|c|}{segmentation example: Hi[0] XXX[0] ,[0] its[0] good[0] to[0] hear[0] from[0] you[0] .[1] it[0] sounds[0] like[0]...} \\ \hline
  \end{tabular}
\end{table}

\subsection*{\textit{Features}}
We utilized three types of features in conjunction with CRF, MLP, BRNN and CRNN: word embeddings as lexical features, punctuation and POS features. Syntactic abstractions of individual words, such as POS tags, have been previously shown to be effective features for similar natural language processing tasks.\cite{liu2005using,treviso2017sentence} To extract POS features, we pre-processed e-Coaching emails using the NLTK POS tagger\footnote{\url{https://www.nltk.org/}}. Punctuation marks, which correspond to one of the symbols \{`.', `,', `!', `?', `:', `;'\} between a pair of words, were also used as a feature, since punctuation marks designate the boundary of a sentence, clause or a phrase and often also correspond to a segment boundary.\cite{cho2002text} For natural language processing (NLP) tasks, inputs are received as textual fragments, in which individual words are as the basic lexico-semantic units. Therefore, it is important to represent a word in such a way that preserves all relevant lexical and semantic information. Embedding is a form of distributed representation, when each word is associated with a dense real-valued vector in low-dimensional space. Embeddings have been previously shown to effectively capture semantic, syntactic and morphological properties of words.\cite{pennington2014glove, mikolov2013distributed} For experiments reported in this paper, we utilized word embeddings pre-trained on Google News corpus consisting of 1.6 billion words using word2vec software package.\footnote{\label{fn:word2vec}\url{https://code.google.com/p/word2vec/}} For words or punctuation marks, which do not have pre-trained embeddings, we utilized the embeddings of the same dimensionality trained on the experimental dataset. CRF utilized lexical features, POS tags and the preceding label.

@@

\textbf{Conditional Random Fields (CRF)}: CRF has been widely used in various NLP tasks that involve sequence annotation, such as part-of-speech tagging.\cite{lafferty2001conditional, hirohata2008identifying} Unlike the maximum-entropy Markov model, which uses per-state exponential models for conditional probability of the next state given a current state, CRF model directly estimates a distribution of the entire output sequence conditioned on the observation sequence. A linear-chain CRF model is defined as a conditional probability distribution $p(y|x)$ of output sequence $y$, given input sequence $x$:

@@

where $Z_x$ is a normalization factor, $f_k(y_{t-1}, y_t, x, t)$ is a feature function, and $\lambda_k$ is a learned weight associated with feature $f_k$. The optimal output sequence $y^*$ for input sequence $x$, $y^* = {arg\,max}_y p(y|x)$, is obtained efficiently using the Viterbi algorithm. In our experiments, the following features were utilized in conjunction with CRF: i) current word or punctuation ii) next and previous 3 words or punctuations iii) binary feature indicating whether a word or punctuation is a special character (';', '?', '.', ',', '!', ':', etc.) or not iv) binary feature indicating whether a word is a title word or not (e.g. ``The'' is a title word but ``the'' is not) v) POS tags.

\textbf{Multi-Layer Perceptron (MLP)}: MLP is a neural network, which consists of multiple fully connected layers that map an input to one or several outputs.\cite{rumelhart1986learning} Figure~\ref{fig:mlp} illustrates a multi-layer perceptron with a single hidden layer. MLPs have no cycles or loops. Information in them flows only forward, from the input layer through the hidden layer(s) to the output layer. The MLP in this study utilizes one hidden layer consisting of 128 neurons and rectified linear unit (ReLU) as a nonlinear activation function. In order to prevent over-fitting, we applied dropout (random masking of neurons) \cite{srivastava2014dropout} to fully connected layers during training. Dropout was also applied to a fully connected layer in CRNN.

@@

\textbf{Bi-directional Recurrent Neural Network (BRNN)}: BRNN is a neural network designed to capture sequential patterns by considering both past  and future inputs as well as complex relationships between input features and output labels.\cite{schuster1997bidirectional} The hidden state of BRNN is an aggregation of the hidden states of a forward and backward recurrent neural networks (RNNs). Gated Recurrent Units (GRU)\cite{chung2014empirical} capable of handling variable size input sequence and having internal memory, which can be reset, were utilized as an RNN in this work.

\textbf{Convolutional Recurrent Neural Network (CRNN)}: CRNN \cite{treviso2017sentence} shown in Figure~\ref{fig:crnn} is a deep neural architecture, which combines convolutional and recurrent layers. Our implementation of CRNN consists of 5 layers: 1) input layer 2) embedding layer 3) convolution layer with max pooling 4) BRNN layer 5) fully connected layer with dropout and sigmoid output. E-coaching email exchanges are represented as a sequence of $m$ words and punctuations, which are fed into the input and embedding layers to produce a $m \times n_e$ matrix after fetching the embeddings for words and punctuations in the input sequence. This matrix is a distributed representation of an input email exchange, which contains rich morpho-syntactic information that can be utilized for its segmentation. When POS tags are utilized along with word embeddings, they are represented with a 10-dimensional vector, which is concatenated with 300-dimensional word embeddings to obtain new embedding vectors $n_e = [n_w;n_p]$ of size 310. The primary purpose of a convolution layer is to extract new features for each word or punctuation mark based on the neighboring words or punctuation marks. A one-dimensional (1D) convolution operation is utilized in this layer in our implementation of BRNN. In 1D convolution, one filter is responsible for extraction of one feature. After applying $n_f$ different filters with zero-padding on both sides of the input text, $n_f$ features are produced by the convolution layer for each word. A max pooling over time operation is then applied to find the most significant features in a textual fragment. The bi-directional recurrent layer receives new features extracted by the convolution layer. Unidirectional RNNs are typically utilized to capture long-range dependencies in a sequence of observations. Bi-directional RNNs, on the other hand, are capable of capturing both past and future contexts through forward and backward traversals of a sequence. The purpose of the fully connected layer in CRNN is to use the output of the bidirectional RNN layer for classifying each word or punctuation into ``new segment'' or ``same segment'' classes. Since a fully connected layer has a larger number of parameters, they are more likely to excessively co-adapt to other parameters in the network and result in over-fitting. To prevent this, we utilized dropout by randomly ignoring 50\% of the connections in the fully connected layer of CRNN. Finally, logistic sigmoid outputs the probability of classifying or labeling each word or punctuation mark with ``same segment'' class. We experimentally determined the optimal parameters using 5-fold cross-validation and found out that the best performance is achieved when filter length in the convolution layer is 7, number of filters is 100, max pooling size is 3, ReLU is used as an activation function in the convolution layer, hyperbolic tangent is used as an activation function in the bi-directional RNN layer and the number of dimensions in the hidden state of RNNs is 200. Adam\cite{kingma2014adam} with 50 epochs, the batch size of 32 and learning rate of 0.001 was used for optimization and the early stopping strategy was applied.\footnote{source code of all methods is available at \url{https://github.com/teanalab/eCoaching-Text-Segmentation}}

\begin{figure}[!htb]
    \centering
    \includegraphics[width=0.95\textwidth]{figures/CRNN.eps}
    \caption{\textbf{Architecture of a convolutional recurrent neural network for automated segmentation of e-Coaching emails into fragments corresponding to MI behaviors}}
    \label{fig:crnn}
\end{figure}

\subsection*{\textit{Evaluation metrics}}
We report standard metrics of precision, recall and F1-measure to evaluate the performance of the classifiers.\cite{aas1999text} Accuracy is not reported as a performance metric, since it is highly sensitive to the distribution of prior class probabilities, which is skewed when datasets with unbalanced classes are involved. The results are reported based on 5-fold cross-validation (one fold was used as a test set and the remaining 4 folds were used as a training set) and weighted macro-averaging over the folds. We also report the area under the precision-recall curve (AUPR), due to its effectiveness in measuring the performance of binary classifiers in the case of the datasets with imbalanced class distribution.\cite{davis2006relationship}

\section*{Results}
Our experiments spanned three dimensions. First, we determined the optimal sizes of word embedding vectors and the sliding window of MLP. Second, we evaluated the performance of different methods with respect to detecting ``new segment'' as well as the weighted average over ``new segment'' and ``same segment'' classes. Third, we assessed the impact of different types of features as well as their combination on the performance of different machine learning methods on the e-Coaching email segmentation task.
